---
title: "Endogeneous infant mortality"
author: "Göran Broström"
date: "January 6, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is convenient to transform data according to the "log-cube transformation"

$$
g(T) = \log^3(1 + 365 T), \quad 0 \le T \le 1,
$$
before fitting survival models, because on that scale, the *BB* model prescribes the *exponential* distribution. Furthermore, we are interested in splitting the first year of life into two periods, the first month and the last eleven. On the new time scale, the cut point is $\log^3(1 + 30)$ =  `r round((log(1 + 30))^3, 1)`.

```{r transform}
g <- function(t) (log(1 + 365 * t))^3 # the log-cube transform 
library(skum)
boys <- obs[obs$birthdate > 1832 & obs$birthdate < 1890 & obs$sex == "male", ]
boys$event <- boys$sluttyp == 2 # death
boys <- boys[, c("id", "region", "birthdate", "enter", "exit", "event")]
boys <- age.window(boys, c(0, 1)) # first year of life
boys$enter <- g(boys$enter) # transform 'enter'
boys$exit <- g(boys$exit) # transform 'exit'
boys1 <- age.window(boys, c(0, g(1/12))) # first month
boys2 <- age.window(boys, c(g(1/12), g(1))) # last 11 months
```

## The Bengtsson-Broström model

So let us check the *Bengtsson-Broström* (BB) model on *boys2*, the last eleven months.

```{r checkbb2}
fitw <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "weibull")
fitc <- coxreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region,
               data = boys2)
check.dist(fitc, fitw, col = c("blue", "red"))
```

An extraordinary excellent fit to the *Weibull* distribution! But the BB model prescribes an exponential distribution, which corresponds to a Weibull with *shape = 1*. So let us look at the parameter estimates.

```{r}
summary(fitw)
```

the shape parameter is estimated to `r exp(fitw$coef[4])`, which is as close to one as you can get. A complete success for the *BB* model.

Now let us look at the first month of life:

```{r checkbb1}
fitw1 <- phreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
               data = boys1, dist = "weibull")
fitc1 <- coxreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
                data = boys1)
check.dist(fitc1, fitw1, col = c("blue", "red"))
```

Not so convincing, (i) the fit to a Weibull is slightly worse, and (2) the cumulative hazards are not straight lines, implying deviation from the exponential distribution.

## The Bourgeouis-Pichat suggestion about exogeneity

But this is what Bourgeouis-Pichat is predicting: During months 2--12 mortality is strictly exogeneous, and we can expect a uniform (BB: exponential) distribution, while during the first month there are both exogeneous and endogeneous causes of death, and deviation from the exponential distribution is to be expected.

The estimated coefficients are

```{r resend}
summary(fitw1)
```

an we can see that the *shape* parameter is significantly different from one.

One way, the PB way(?), to estimate the contribution of the endogeneous deaths is to assume that the exogeneous mortality is exponential during all twelve months, and extrapolate.

```{r extrapo, fig.cap = "Skellefteå and Umeå boys, 1820-1950."}
fitex <- phreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
               data = boys2, dist = "weibull")
fitee <- coxreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
                data = boys1)
plot(fitex, xlim = c(0, g(1/12)), col = "blue", fn = "cum", 
     ylim = c(0, 0.03), main = "Neonatal mortality")
y <- c(0, cumsum(fitee$hazards[[1]][, 2]))
x <- c(0, fitee$hazards[[1]][, 1])
lines(x, y, col = "red")
abline(h = 0)
text(10, 0.020, "Total mortality", col = "red")
text(25, 0.012, "Exogeneous mortality", col = "blue")
```

Obviously the endogeneous mortality dominates the first month of life. (We are still in the transformed time scale, with "40" representing 1 month.)

## Without the transformation

We repeat everything, but without transforming.

```{r transform2}
g <- function(t) t # the identity transform! 
library(skum)
boys <- obs[obs$birthdate > 1820 & obs$sex == "male", ]
boys$event <- boys$sluttyp == 2 # death
boys <- boys[, c("id", "region", "birthdate", "enter", "exit", "event")]
boys <- age.window(boys, c(0, 1)) # first year of life
boys$enter <- g(boys$enter) # transform 'enter'
boys$exit <- g(boys$exit) # transform 'exit'
boys1 <- age.window(boys, c(0, g(1/12))) # first month
boys2 <- age.window(boys, c(g(1/12), g(1))) # last 11 months
```

### The Bengtsson-Broström model, II

So let us check the *Bengtsson-Broström* (BB) model on *boys2*, the last eleven months.

```{r checkbb22}
fitw <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "weibull")
fitc <- coxreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region,
               data = boys2)
check.dist(fitc, fitw, col = c("blue", "red"))
```

Not an equally excellent fit to the *Weibull* distribution! 

And, add (12 jan 2020) an experiment with the Gompertz distribution:

```{r gomp}
fitg <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "gompertz", param = "rate")
check.dist(fitc, fitg, col = c("blue", "red"), main = "Gompertz")
```


## Conclusion

Maybe it is too early to completely reject the log-cube transformation (and the BB model) for infant mortality.