---
title: "Create survival data frame for BP"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
#bibliography: snus2.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
- \subsectionfont{\normalfont\itshape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
requireNamespace("eha")
```

# Introduction

We are reading data from *Uttag U22003* from Cedar, containing infant mortality 
data from Västerbotten. The original is stored in *RDS* format in the folder *Data*, 
two files *individ* and *boende*.

The goal is th create one file suitable for *survival analysis*, meaning that the
building blocks for *episodes* are there in the form of three variables named *enter*
(start age of episode), *exit* (end age of episode), and *event* (indicator of death).
Key covariates are further *birthdate* and *place of birth*.

*Revision December 5, 2022:* Only infants with known birthdate to start with. Infants with incomplete death date are includeed as survivors, provided that it can be established that the 
infant survived one year of age.

# The individual file

Let's call it *infant*. 


*NOTE:* Skip this and concentrate on *boende* for intervals.

We first separate the data into *survivors* **(is.na(deathdate))** and
*deaths* **(!is.na(deathdate))**, ad then we deal with the two groups separately.

Finally, we merge the two sets.

*END skip*

```{r read, individ}
ind <- readRDS("../Bio/Data/individ.rds")
infant <- data.frame(region = ind$REGPREFIX, 
                     fodfrs = ind$FODHEMFRS,
                     id = ind$UTTAGSID, 
                     sex = ind$KON, 
                     birthdate = ind$FODDAT_U, 
                     bdupd = ind$FODDAT_IND, 
                     bapdate = ind$DOPDAT_U, 
                     bapupd = ind$DOPDAT_IND, 
                     ab = ind$AB,
                     fb = ind$FB,
                     deathdate = ind$DODDAT_U, 
                     ddupd = ind$DODDAT_IND, 
                     cause1 = substr(ind$ICD10H_1, 1, 1), ## NOTE: Added 2023-02-09
                     cause2 = substr(ind$ICD10H_2, 1, 1), ## NOTE: Added 2023-02-09
                     cause3 = substr(ind$ICD10H_3, 1, 1), ## NOTE: Added 2023-02-09
                     cause4 = substr(ind$ICD10H_4, 1, 1), ## NOTE: Added 2023-02-09
                     cause5 = substr(ind$ICD10H_5, 1, 1), ## NOTE: Added 2023-02-09
                     cause6 = substr(ind$ICD10H_6, 1, 1), ## NOTE: Added 2022-02-09
                     outdate = ind$BOSDAT_U,
                     odupd = ind$BOSDAT_IND,
                     uttyp = ind$BOSTYP, 
                     parish = ind$FODHEMFRS,
                     m_parish = ind$MORBOORTKOD, 
                     hisco = ind$FAR_HISCOKOD_1, 
                     cause = ind$DODORSSTD_1,
                     icd10h = ind$ICD10H_1)
inf <- infant # Save a copy of the original.
infant <- infant[infant$bdupd < 0.5, ] # Known birthdate.
infant <- infant[infant$sex %in% 1:2, ] # Known sex
infant$sex <- factor(infant$sex, labels = c("boy", "girl"))
#
infant$birthdate <- as.Date(as.character(infant$birthdate), 
                            format = "%Y%m%d")
infant$bapdate <- as.Date(as.character(infant$bapdate), 
                            format = "%Y%m%d")
infant$deathdate <- as.Date(as.character(infant$deathdate), 
                            format = "%Y%m%d")
infant$outdate <- as.Date(as.character(infant$outdate), 
                          format = "%Y%m%d")
survs <- infant[is.na(infant$deathdate), ]
deaths <- infant[!is.na(infant$deathdate), ]
```


## Survivors

"Survivors" means here individuals with unknown death date.

We need to find a date *last seen* and the corresponding *exposure time*.

```{r survivors}
exp.time <- with(survs, as.numeric(outdate - birthdate))

survs$exit <- pmin(exp.time, 365) # Note 'p'!
survs <- survs[survs$exit > 0.5, ]
survs$event <- 0
survs$enter <- 0
##summary(exp.time)
```

## Deaths

```{r deaths}
exp.time <- with(deaths, as.numeric(deathdate - birthdate))
deaths$exit <- exp.time
deaths <- deaths[deaths$exit >= -0.0001, ] # Safety?!.....
deaths$exit[deaths$exit <= 0.01] <- 0.25 # Dies on birth day: Six hours of life...
deaths <- deaths[(deaths$ddupd < 0.5) | (deaths$exit > 365), ]
deaths$event <- 1
deaths$enter <- 0
deaths <- eha::age.window(deaths, c(0, 365))
```

## Put together and save

```{r together}
infant <- rbind(deaths, survs)
infant <- infant[order(infant$birthdate), ]
infant$icd.chapter <- substr(infant$icd10h, 1, 1)
is.na(infant$icd.chapter) <- infant$icd.chapter == ""
infant$icd.chapter <- factor(infant$icd.chapter)
infant <- infant[infant$birthdate > "1860-12-31", ]
infant <- infant[infant$birthdate < "1951-01-01", ]
infant$socst <- (infant$hisco %/% 10000) %in% 1:4
infant$socst <- factor(infant$socst, labels = c("Other", "Elite"))
saveRDS(infant, file = "Rdata/infant.rds")
summary(infant)
```
